\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-12 paper/ms.tex
a few edits

\response{Existing astronomical time series datasets have already reached the
limit where na\"ive application GP models is no longer tractable.
NASA's \kepler\ Mission \citep{Borucki:2010}, for example, measured light
curves with more than 60,000 observations each for about 190,000 stars.
Current and forthcoming surveys such as \project{K2} \citep{Howell:2014},
\tess\ \citep{Ricker:2014}, \lsst\ \citep{Ivezic:2008}, \project{WFIRST}
\citep{Spergel:2015}, and PLATO \citep{Rauer:2014} will continue to produce
similar or larger data volumes.}

@@

Another benefit of \celerite\ compared to Kalman filter solvers is the fact
that it is a direct solver for the relevant matrix equations instead of just
an algorithm to compute \eq{gp-likelihood}.
This means that the inverse covariance matrix $K^{-1}$ can be applied to
general vectors and matrices in $\mathcal{O}(N)$, while reusing the
factorization.
This is a crucial feature when GP models are combined with linear regression
\citep[for example][]{Luger:2017}.
}

@@

%As discussed previously, the \celerite\ method is based upon a method
%developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
%In the case where the kernel function has a single term of the form given in
%\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
%and the Rybicki \& Press method.
%Their method does not require solving the extended matrix because the matrix
%inverse can be calculated analytically, but the method is made slightly more
%complicated when white noise is included.
%We implement their full method and find empirically that it is more
%computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
%but we note that the requirement of a single, real, exponential kernel is
%extremely restrictive and this model cannot describe quasi-periodic
%variability.

Another popular method uses the fact that, in the limit of evenly-spaced data
and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
\citep[for example][]{Dillon:2013}.
There are exact methods for solving Toeplitz matrix equations that scale as
$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
\citep{Wilson:2014}.
The Toeplitz method is, in some ways, more flexible than \celerite\ because it
can be used with any stationary kernel, but it requires uniformly spaced data
and the scaling is worse than \celerite\ so it is less efficient when applied
to large datasets.

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key takeaway here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific, but flexible, class of one-dimensional kernel
functions.
Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
stochastically-driven, damped simple harmonic oscillators, they are a
physically motivated choice of covariance function in many astronomical applications.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis, but the fact that the computational cost generally
scales as the cube of the number of data points has limited their use to relatively
small datasets.
With the linear scaling of \celerite\, we envision the application of Gaussian
processes may be expanded to much larger datasets.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped,
stochastically-driven harmonic oscillator matches simple models of
astrophysical variability, and makes the parameterization interpretable in
terms of resonant frequency, amplitude, and quality factor.
A drawback of this method is its quadratic scaling with the number of
terms $J$, but, in some cases, small values of $J$ may suffice.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar light curves when detecting or characterizing transiting planets
\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
All of these analyses have been limited to small datasets or restrictive
kernel choices.
\celerite\ weakens these requirements by providing a scalable method for
computing the likelihood and a physical motivation for the choice of kernel.
\celerite\ can be used to model stellar oscillations using the relation to the
mixture of stochastically-driven, damped simple harmonic oscillators, as
we have demonstrated in this paper.
As higher signal-to-noise observations of transiting exoplanet systems are
obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters.
We expect that \celerite\ will be important for transit detection
\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
functions \citep{Knutson:2007, Zucker:2007}, and more.

@@

The \celeriteterm\ formalism can also be used for power spectrum estimation and
quantification of its uncertainties.
In principle, a large number of \celeriteterm\ terms could be used to perform
non-parametric probabilistic inference of the power spectrum despite
unevenly-spaced data with heteroscedastic noise \citep[for
example,][]{Wilson:2013, Kelly:2014}.
This type of analysis will be limited by the quadratic scaling of \celerite\
with the number of terms $J$, but this limits existing methods as well
\citep[CARMA models,][]{Kelly:2014}.
In general, we encourage the use of physically motivated models for parameter
estimation instead of qualitative modeling of the power spectrum itself.

There are many data analysis problems where \celerite\ will not be applicable.
In particular, the restriction to one-dimensional problems is significant.
There are many examples of multidimensional GP modeling the astrophysics
literature \citep[recent examples from the field of exoplanet characterization
include][]{Haywood:2014, Rajpaul:2015, Aigrain:2016}, where \celerite\ cannot
be used to accelerate any of these analyses.
It is plausible that an extension could be derived to tackle some
multidimensional problems with the right structure~--~simultaneous parallel
time series, for example~--~and we hope to revisit this possibility in future
work.

@@

The product of a rank-$R$ semiseparable matrix with general vectors and
matrices can be computed in $\mathcal{O}(N\,R)$ operations
\citep{Vandebril:2007}.
In the context of \celeriteterm\ models, na\"ive application of these methods
will result in numerical overflow and underflow issues much like we found with
the Cholesky factorization.
To avoid these errors, we derive the following numerically stable algorithm
for computing the dot product of a \celeriteterm\ covariance matrix
\begin{eqnarray}
\bvec{y} = K\,\bvec{z}
\end{eqnarray}
using the semiseparable representation for $K$ defined in \eqalt{reparam}:
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j}\,\left[f^+_{n+1,j} +
    \tilde{U}_{n,j}\,z_{n+1}\right] \\
f^-_{n,j} &=& \phi_{n,j}\,\left[f^-_{n-1,j} +
    \tilde{V}_{n-1,j}\,z_{n-1}\right] \\
y_n &=& A_{n,n}\,z_n + \sum_{j=1}^{2\,J} \left[ \tilde{V}_{n,j} f^+_{n,j} +
    \tilde U_{n-1,j} f^-_{n,j}\right] \quad,
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$.
This algorithm requires two sweeps~--~$n=2,\,\ldots,\,N$ to compute $f^-$ and
$n=N-1,\,\ldots,\,1$ to compute $f^+$~--~with a scaling of $\mathcal{O}(N\,J)$.

>>>>

\response{Existing astronomical time series datasets have already reached the
limit where na\"ive application GP models is no longer tractable.
NASA's \kepler\ Mission \citep{Borucki:2010}, for example, measured light
curves with more than 60,000 observations each for about 190,000 stars.
Current and forthcoming surveys such as \project{K2} \citep{Howell:2014},
\tess\ \citep{Ricker:2014}, \lsst\ \citep{Ivezic:2008}, \project{WFIRST}
\citep{Spergel:2015}, and \project{PLATO} \citep{Rauer:2014} will continue to
produce similar or larger data volumes.}

@@

Another benefit of \celerite\ compared to Kalman filter solvers is the fact
that it is an algebraic solver for the relevant matrix equations instead of
an algorithm to directly compute \eq{gp-likelihood}.
This means that the inverse covariance matrix $K^{-1}$ can be applied to
general vectors and matrices in $\mathcal{O}(N)$, while reusing the
factorization.
This is a crucial feature when GP models are combined with linear regression
\citep[for example][]{Luger:2017}.
}

@@

Another popular GP method uses the fact that, in the limit of evenly-spaced
data and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
\citep[for example][]{Dillon:2013}.
There are exact methods for solving Toeplitz matrix equations that scale as
$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
\citep{Wilson:2014}.
The Toeplitz method is, in some ways, more flexible than \celerite\ because it
can be used with any stationary kernel, but it requires uniformly spaced data
and the scaling is worse than \celerite\ so it is less efficient when applied
to large datasets.

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key takeaway here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific class of one-dimensional kernel functions.
Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
stochastically-driven, damped simple harmonic oscillators, they are a
physically motivated choice of covariance function in many astronomical applications.

@@

Gaussian Process models have been fruitfully applied to many problems in
astronomical data analysis, but the fact that the computational cost generally
scales as the cube of the number of data points has limited their use to
small datasets with $N \lesssim 1000$.
With the linear scaling of \celerite\, we envision that the application of
Gaussian processes will be expanded to the existing and forthcoming large
astronomical time domain surveys such as \kepler\ \citep{Borucki:2010},
\project{K2} \citep{Howell:2014}, \tess\ \citep{Ricker:2014}, \lsst\
\citep{Ivezic:2008}, \project{WFIRST} \citep{Spergel:2015}, and
\project{PLATO} \citep{Rauer:2014}.
Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
number of components it is flexible enough to describe a wide range of
astrophysical variability.
In fact, the relation of the \celeriteterm\ kernel to the damped,
stochastically-driven harmonic oscillator matches simple models of
astrophysical variability, and makes the parameterization interpretable in
terms of resonant frequency, amplitude, and quality factor.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt GP methods for analyzing the noise in stellar light
curves and radial velocity datasets when detecting or characterizing
transiting planets \citep[for example,][]{Carter:2009, Gibson:2012,
Haywood:2014, Barclay:2015, Evans:2015, Rajpaul:2015, Aigrain:2016,
Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
All of these analyses have been limited to small datasets or restrictive
kernel choices, but \celerite\ weakens these requirements by providing a
scalable method for computing the likelihood and a physical motivation for the
choice of kernel.
As higher signal-to-noise observations of transiting exoplanet systems are
obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters.
We predict that \celerite\ will be important for scaling methods of transit
detection \citep{Pope:2016, Foreman-Mackey:2016b}, transit timing
\citep{Agol:2005, Holman:2005}, transit spectroscopy \citep{Brown:2001},
Doppler beaming \citep{Loeb:2003, Zucker:2007}, tidal distortion
\citep{Zucker:2007}, and phase functions \citep{Knutson:2007, Zucker:2007} to
the large astronomical time domain surveys of the future.

@@

The \celeriteterm\ formalism can also be used for power spectrum estimation and
quantification of its uncertainties.
A mixture of \celeriteterm\ terms can be used to perform non-parametric
probabilistic inference of the power spectrum despite unevenly-spaced data
with heteroscedastic noise \citep[see][for examples of this
procedure]{Wilson:2013, Kelly:2014}.
This type of analysis will be limited by the quadratic scaling of \celerite\
with the number of terms $J$, but this limits existing methods as well
\citep{Kelly:2014}.

There are many data analysis problems where \celerite\ will not be immediately
applicable.
In particular, the restriction to one-dimensional problems is significant.
There are many examples of multidimensional GP modeling the astrophysics
literature \citep[recent examples from the field of exoplanet characterization
include][]{Haywood:2014, Rajpaul:2015, Aigrain:2016}, where \celerite\ cannot
be used to speed up the analysis.
It is plausible that an extension can be derived to tackle some
multidimensional problems with the right structure~--~simultaneous parallel
time series, for example~--~and we hope to revisit this possibility in future
work.

@@

The product of a rank-$R$ semiseparable matrix with general vectors and
matrices can be computed in $\mathcal{O}(N\,R)$ operations
\citep{Vandebril:2007}.
In the context of \celeriteterm\ models, na\"ive application of these methods
will result in numerical overflow and underflow issues, much like we found
with the Cholesky factorization.
To avoid these errors, we derive the following numerically stable algorithm
for computing the dot product of a \celeriteterm\ covariance matrix
\begin{eqnarray}
\bvec{y} = K\,\bvec{z}
\end{eqnarray}
using the semiseparable representation for $K$ defined in \eqalt{reparam}:
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j}\,\left[f^+_{n+1,j} +
    \tilde{U}_{n,j}\,z_{n+1}\right] \\
f^-_{n,j} &=& \phi_{n,j}\,\left[f^-_{n-1,j} +
    \tilde{V}_{n-1,j}\,z_{n-1}\right] \\
y_n &=& A_{n,n}\,z_n + \sum_{j=1}^{2\,J} \left[ \tilde{V}_{n,j} f^+_{n,j} +
    \tilde U_{n-1,j} f^-_{n,j}\right] \quad,
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$.
This algorithm requires two sweeps~--~$n=2,\,\ldots,\,N$ to compute $f^-$ and
$n=N-1,\,\ldots,\,1$ to compute $f^+$~--~with a scaling of $\mathcal{O}(N\,J)$.
